---
title: |
  |Principal Component Analysis
author: "Sajia Darwish"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  html_document:
    toc: yes
    toc_depth: 4
  pdf_document:
    toc: yes
    toc_depth: '4'
keep_tex: yes
theme: bclear
fontsize: 10pt
geometry: margin=1in
header-includes:
- \usepackage{enumitem}
- \usepackage{indentfirst}
- \usepackage[default, scale=1]{raleway}
- \usepackage[T1]{fontenc}
- \usepackage{inconsolata}
always_allow_html: yes
---

```{r, echo=FALSE}
library(tidyverse)
#install.packages("rlang")
library(tidymodels)
library(corrplot)
library(ggrepel)
library(tidytext)
library(dotwhisker) # for coefficient plots

# hmo <- read.csv("/data/KI/imic/data/raw_lab_data/elicit/milk_analytes/HMO_ELICIT.csv")
# hmo <- rename("bmid" = "X", hmo)
# 
# sapient <- read.csv("/data/KI/imic/data/raw_lab_data/elicit/milk_analytes/Sapient_ELICIT.csv")
# sapient <- rename("bmid_base" = "X", sapient)
# 
# bvit <- read.csv("/data/KI/imic/data/raw_lab_data/elicit/milk_analytes/Allen_Bvit_ELICIT.csv")
# bvit <- rename("bmid_base" = "X", bvit)
# 
biocratesNorm <- read.csv("/data/KI/imic/data/raw_lab_data/elicit/milk_analytes/Biocrates_ELICIT_Normalized_toALL_NS.csv")
biocratesNorm <- rename("bmid_base" = "X", biocratesNorm)

# biocrates <- read.csv("/data/KI/imic/data/raw_lab_data/elicit/milk_analytes/Biocrates_ELICIT_NOT_Normalized.csv")
# biocrates <- rename("bmid_base" = "X", biocrates)
# 
# metabolInd <- read.csv("/data/KI/imic/data/raw_lab_data/elicit/milk_analytes/MetaboIndicators/Biocrates_ELICIT_Normalized_MetaboINDICATOR.csv")
# metabolInd <- rename("bmid_base" = "X", metabolInd)
```

# There are a few steps to take before PCA:
1) Replace < LOD with with the minimum observed value for each variable divided by 2.
   Note: after this calculation, some values go to infinity. They have been set to 0.
2) Check missingness in the data: after cleaning = no missingness.
3) Remove zero variance variables.
4) Scale and center all variables to ensure that the criterion for finding linear combinations of the predictors is based on how much variation they explain and therefore improve numerical stability.

```{r echo=FALSE, message=FALSE, warning=FALSE}
data <- biocratesNorm %>%
  select(-bmid_base)

# Change <LOD to NA.
for (i in 1:ncol(data)) {
  data[, i] = ifelse(data[, i] == "< LOD", NA, data[, i])
}

# Make all variables numeric.
#str(data)
data <- as.data.frame(sapply(data, as.numeric))

# Prepare for test beforehand
#min(data $C5.OH..C3.DC.M., na.rm = T) / 2 #0.040535 #Correct

# Change these values to min/2.
data <- data %>% 
  mutate_if(is.numeric, function(x) ifelse(is.na(x), min(x, na.rm = T) / 2, x))

# Change Inf values to 0.
for (i in 1:ncol(data)) {
  data[, i] = ifelse(data[, i] == "Inf", 0, data[, i])
}

# Check for missingness.
#table(is.na(data))

# Remove columns with no variance.
data <- data[ - as.numeric(which(apply(data, 2, var) == 0))]
#str(data)
```

# PCA using the normalized ELICIT targeted metabolomics data

## Inspect zoomed-in scree plots after running PCA
```{r, echo=FALSE}
pca = prcomp(data, center = TRUE, scale = TRUE)

# Variability of each principal component: pr.var
pr.var <- pca $ sdev ^ 2

# Variance explained by each principal component: pve
pve <- pr.var / sum(pr.var)

# # Plot variance explained for each principal component
# plot(pve, xlab = "Principal Component",
#      ylab = "Proportion of Variance Explained",
#      ylim = c(0, 1), type = "b")

# # Plot cumulative proportion of variance explained
# plot(cumsum(pve), xlab = "Principal Component",
#      ylab = "Cumulative Proportion of Variance Explained",
#      ylim = c(0, 1), type = "b")

# Let's zoom in a bit - nope, a lot.
plot(pve, xlab = "Principal Component",
     ylab = "Proportion of Variance Explained",
     ylim = c(0, 1), xlim = c(0, 25), type = "b") # 6-10 components = optimal.

# Let's zoom in by a lot in here as well
plot(cumsum(pve), xlab = "Principal Component",
     ylab = "Cumulative Proportion of Variance Explained",
     ylim = c(0, 1), xlim = c(0, 25), type = "b")
```

# Examine the PCs
```{r}
# Create a recipe
char_recipe <- recipe(~ ., data = data) %>%
  # Tell PCA that these are the identification variables and to not use them.
  #update_role(bmid_base, new_role = "id") %>%
  # Scale so the range of variables don't affect our analysis.
  step_normalize(all_predictors()) %>%
  step_pca(all_predictors(), num_comp = 10) # Useful for juice().

# Prep the recipe and take a look at it. 
pca_estimates_prep <- prep(char_recipe)

# Extract pca results: loadings
pca_estimates <- tidy(pca_estimates_prep, 2)

# Get data in shape to plot
pca_estimates2 <- pca_estimates %>%
  filter(component %in% paste0("PC", 1:10)) %>%
  mutate(importance = value * value,
         component = fct_inorder(component),
         terms = reorder_within(terms, # order within component.
                                by = importance,
                                within = component)) %>%
  # Get the highest 10 contributors by importance for each component
  group_by(component) %>%
  slice_max(importance, n = 10)

# Plot the highest 10 contributors to each component by importance
pca_estimates2 %>%
  ggplot() +
  geom_col(aes(y = importance, 
               x = terms, 
               fill = importance < 0.002)) + # 1/603 (number of PCA variables)
  geom_hline(yintercept = 0.002) +
  facet_wrap(~ component, 
             nrow = 2,
             scales = "free") + 
  guides(fill = "none") +
  scale_x_reordered() +
  xlab("") + ylab("") +
  coord_flip() +
  ggtitle("10 top contributors to each component by importance") +
  theme(axis.text.x = element_blank(),
        axis.ticks = element_blank())
```

```{r}
# Plot the highest 10 contributors to each component by value to get direction
pca_estimates2 %>%
  ggplot() +
  geom_col(aes(y = value, 
               x = terms, 
               fill = importance < 0.002)) +
  geom_hline(yintercept = 0.002) +
  facet_wrap(~ component, 
             nrow = 2,
             scales = "free") + 
  guides(fill = FALSE) +
  scale_x_reordered() +
  xlab("") + ylab("") +
  coord_flip() +
  ggtitle("10 top contributors to each component by value") +
  theme(axis.text.x = element_blank(),
        axis.ticks = element_blank())
```

# Merge the PC data with ID's and save dataset.
```{r}
dataPC <- juice(pca_estimates_prep)
bmid_base = biocratesNorm[, 1]
fullData <- cbind(bmid_base, dataPC)
```















