---
title: |
  TMLE Results
author: "Sajia Darwish"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  pdf_document:
    toc: yes
    toc_depth: '4'
  html_document:
    toc: yes
    toc_depth: 4
keep_tex: yes
theme: bclear
fontsize: 10pt
geometry: margin=1in
header-includes:
- \usepackage{enumitem}
- \usepackage{indentfirst}
- \usepackage[default, scale=1]{raleway}
- \usepackage[T1]{fontenc}
- \usepackage{inconsolata}
always_allow_html: yes
---

# Overview

- Implemented TMLE over each group of biomarkers separately for each study, controlling for only a couple of baseline covariates for simplicity (sex and mother's age for most).
- The following graphs show TMLE estimates for each group of biomarkers to understand how they are associated with the outcome (haz_6m).
- The outcome for these analyses were simualted from the distrubutions of the actual outcomes.

```{r, include=FALSE}
library(tidyverse)

hmo <- read.csv("/data/imic/results/tmle/hmoElicitTMLE.csv")
bioc <- read.csv("/data/imic/results/tmle/biocElicitTMLE.csv")
bvit <- read.csv("/data/imic/results/tmle/BvitElicitTMLE.csv")
metabol <- read.csv("/data/imic/results/tmle/MetabolElicitTMLE.csv")
```

# ELICIT

## HMO
```{r, echo=FALSE, fig.show="hold", out.width="75%", fig.align = 'center'}
hmo %>%
    ggplot(aes(x =tmle_est)) +
    geom_histogram(binwidth = 0.1, fill="#69b3a2", color="#e9ecef", alpha = 0.9) +
    theme_classic()
```

## Biocrates
```{r, echo=FALSE, fig.show="hold", out.width="75%", fig.align = 'center'}
bioc %>%
    ggplot(aes(x =tmle_est)) +
    geom_histogram(binwidth = 0.1, fill="#69b3a2", color="#e9ecef", alpha = 0.9) +
    theme_classic()
```

## B-vitamins
```{r, echo=FALSE, fig.show="hold", out.width="75%", fig.align = 'center'}
bvit %>%
    ggplot(aes(x =tmle_est)) +
    geom_histogram(binwidth = 0.1, fill="#69b3a2", color="#e9ecef", alpha = 0.9) +
    theme_classic()
```

## Metabolomcis
```{r, echo=FALSE, fig.show="hold", out.width="75%", fig.align = 'center'}
metabol %>%
    ggplot(aes(x =tmle_est)) +
    geom_histogram(binwidth = 0.1, fill="#69b3a2", color="#e9ecef", alpha = 0.9) +
    theme_classic()
```


















