---
title: |
  TMLE Results
author: "Sajia Darwish"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  pdf_document:
    toc: yes
    toc_depth: '4'
  html_document:
    toc: yes
    toc_depth: 4
keep_tex: yes
theme: bclear
fontsize: 12pt
geometry: margin=1in
header-includes:
- \usepackage{enumitem}
- \usepackage{indentfirst}
- \usepackage[default, scale=1]{raleway}
- \usepackage[T1]{fontenc}
- \usepackage{inconsolata}
always_allow_html: yes
---

# Overview

- Implemented TMLE over each group of biomarkers separately for each study, controlling for only a couple of baseline covariates for simplicity (sex and mother's age for most).
- The following graphs show TMLE estimates for each group of biomarkers to understand how they are associated with the outcome (haz_6m).
- The outcome for these analyses were simualted from the distrubutions of the actual outcomes.

```{r, include=FALSE}
library(tidyverse)

hmo <- read.csv("/data/imic/results/tmle/hmoElicitTMLE.csv")
bioc <- read.csv("/data/imic/results/tmle/biocElicitTMLE.csv")
bvit <- read.csv("/data/imic/results/tmle/BvitElicitTMLE.csv")
metabol <- read.csv("/data/imic/results/tmle/MetabolElicitTMLE.csv")
```

# ELICIT (N = 199)

## HMO (22 features)
```{r, echo=FALSE, fig.show="hold", out.width="75%", fig.align = 'center'}
hmo %>%
    ggplot(aes(x =tmle_est)) +
    geom_histogram(binwidth = 0.1, fill="#69b3a2", color="#e9ecef", alpha = 0.9) +
    theme_classic()
```

# Volcano plot
```{r}
# Calculate p-value
hmo $ z = hmo $ tmle_est / hmo $ se
hmo $ pvalue = exp((-0.717 * hmo $ z) - (0.416 * (hmo $ z^2)))

# Calculate log 2 fold change [TO DO: get unadjusted estimates]
# Name this variable log2FoldChange

# Convert directly in the aes()
p <- ggplot(data = hmo, aes(x = tmle_est, y = log10(pvalue))) + geom_point() +
  theme_minimal()

# Add vertical lines for log2FoldChange thresholds, and one horizontal line for 
# the p-value threshold 
p2 <- p + geom_vline(xintercept = c(-0.6, 0.6), col = "red") +
    geom_hline(yintercept = -log10(0.05), col = "red")

# The significantly different are the ones found in the upper-left and upper-right corners.
# Add a column to the data frame to specify if they are UP- or DOWN- (log2FoldChange respectively positive or negative)

# add a column of NAs
hmo $ state <- "NO"
# if log2Foldchange > 0.6 and pvalue < 0.05, set as "UP" 
hmo $ state[hmo $ log2FoldChange > 0.6 & hmo $ pvalue < 0.05] <- "UP"
# if log2Foldchange < -0.6 and pvalue < 0.05, set as "DOWN"
hmo $ state[hmo $ log2FoldChange < -0.6 & hmo $ pvalue < 0.05] <- "DOWN"

# Re-plot but this time color the points with "state"
p <- ggplot(data = hmo, aes(x = log2FoldChange, y = -log10(pvalue), 
                          col = state)) + geom_point() + theme_minimal()

# Add lines as before...
p2 <- p + geom_vline(xintercept = c(-0.6, 0.6), col = "red") +
        geom_hline(yintercept = -log10(0.05), col = "red")

# Change point color 
mycolors <- c("blue", "red", "black")
names(mycolors) <- c("DOWN", "UP", "NO")
p3 <- p2 + scale_colour_manual(values = mycolors)

# Now write down the names of variables beside the points
# Create a new column "delabel" to de, that will contain the name of genes differentially expressed (NA in case they are not)
hmo $ delabel <- NA
hmo $ delabel[hmo $ status != "NO"] <- hmo $ gene_symbol[hmo $ status != "NO"]

ggplot(data = hmo, aes(x = log2FoldChange, y = -log10(pvalue), col = status, 
                       label = delabel)) + 
  geom_point() + theme_minimal() + geom_text_repel() + 
  scale_color_manual(values = c("blue", "black", "red")) +
  geom_vline(xintercept = c(-0.6, 0.6), col = "red") +
  geom_hline(yintercept = -log10(0.05), col = "red")

```

## Biocrates (329 features)
```{r, echo=FALSE, fig.show="hold", out.width="75%", fig.align = 'center'}
bioc %>%
    ggplot(aes(x =tmle_est)) +
    geom_histogram(binwidth = 0.1, fill="#69b3a2", color="#e9ecef", alpha = 0.9) +
    theme_classic()
```

## B-vitamins (23 features)
```{r, echo=FALSE, fig.show="hold", out.width="75%", fig.align = 'center'}
bvit %>%
    ggplot(aes(x =tmle_est)) +
    geom_histogram(binwidth = 0.1, fill="#69b3a2", color="#e9ecef", alpha = 0.9) +
    theme_classic()
```

## Metabolomcis (111 features)
```{r, echo=FALSE, fig.show="hold", out.width="75%", fig.align = 'center'}
metabol %>%
    ggplot(aes(x =tmle_est)) +
    geom_histogram(binwidth = 0.1, fill="#69b3a2", color="#e9ecef", alpha = 0.9) +
    theme_classic()
```

# VITAL (N = 147)
```{r, include=FALSE}
hmo <- read.csv("/data/imic/results/tmle/hmoVitalTMLE.csv")
bioc <- read.csv("/data/imic/results/tmle/biocVitalTMLE.csv")
metabol <- read.csv("/data/imic/results/tmle/MetabolVitalTMLE.csv")
```

## HMO (22 features)
```{r, echo=FALSE, fig.show="hold", out.width="75%", fig.align = 'center'}
hmo %>%
    ggplot(aes(x =tmle_est)) +
    geom_histogram(binwidth = 0.1, fill="#69b3a2", color="#e9ecef", alpha = 0.9) +
    theme_classic()
```

## Biocrates (630 features)
```{r, echo=FALSE, fig.show="hold", out.width="75%", fig.align = 'center'}
bioc %>%
    ggplot(aes(x =tmle_est)) +
    geom_histogram(binwidth = 0.1, fill="#69b3a2", color="#e9ecef", alpha = 0.9) +
    theme_classic()
```

## Metabolomcis (111 features)
```{r, echo=FALSE, fig.show="hold", out.width="75%", fig.align = 'center'}
metabol %>%
    ggplot(aes(x =tmle_est)) +
    geom_histogram(binwidth = 0.1, fill="#69b3a2", color="#e9ecef", alpha = 0.9) +
    theme_classic()
```
















