---
title: |
  | Biomarker datasets
author: "The Berkeley Team"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  html_document:
    toc: yes
    toc_depth: 4
  pdf_document:
    toc: yes
    toc_depth: '4'
keep_tex: yes
theme: bclear
fontsize: 10pt
geometry: margin=1in
header-includes:
- \usepackage{enumitem}
- \usepackage{indentfirst}
- \usepackage[default, scale=1]{raleway}
- \usepackage[T1]{fontenc}
- \usepackage{inconsolata}
always_allow_html: yes
---

# ELICIT
```{r setup1, echo=FALSE, warning = FALSE, message = FALSE}
BV_dir = "/data/KI/imic/"

library(DataExplorer)
library(tidyverse)
library(table1)
library(data.table)
library(SmartEDA)

hmo <- read.csv("/data/KI/imic/data/raw_lab_data/elicit/milk_analytes/HMO_ELICIT.csv")
sapient <- read.csv("/data/KI/imic/data/raw_lab_data/elicit/milk_analytes/Sapient_ELICIT.csv")
bvit <- read.csv("/data/KI/imic/data/raw_lab_data/elicit/milk_analytes/Allen_Bvit_ELICIT.csv")
biocratesNorm <- read.csv("/data/KI/imic/data/raw_lab_data/elicit/milk_analytes/Biocrates_ELICIT_Normalized_toALL_NS.csv")
biocrates <- read.csv("/data/KI/imic/data/raw_lab_data/elicit/milk_analytes/Biocrates_ELICIT_NOT_Normalized.csv")
metabolInd <- read.csv("/data/KI/imic/data/raw_lab_data/elicit/milk_analytes/MetaboIndicators/Biocrates_ELICIT_Normalized_MetaboINDICATOR.csv")
```

## HMOs
Note: “Secretor”, “Diversity”, “Evenness”, “SUM_nmol/mL”, “SUM_ug/mL”, “Sia_nmol/mL”, and “Fuc_nmol/mL” are summary variables. The suffix “_nmol/mL”, “_ug/mL” and “_pct” indicate different units for the same HMOs.
```{r, echo=FALSE}
# hmo $ X = NULL
# hmoElicit <- table1(~ ., data = hmo)
# saveRDS(hmoElicit, file = paste0(BV_dir, "/results/hmoElicit.RDS"))
# readRDS("/data/KI/imic/results/hmoElicit.RDS")
ExpData(data = hmo, type = 1)
print("Sample data")
head(hmo)[15:17]
print("Sample column names")
names(hmo)[10:50]
```

## B-Vitamins
```{r, echo=FALSE}
# bvit $ X = NULL
# bvitElicit <- table1(~ ., data = bvit)
# saveRDS(bvitElicit, file = paste0(BV_dir, "/results/bvitElicit.RDS"))
# readRDS("/data/KI/imic/results/bvitElicit.RDS")
ExpData(data = bvit, type = 1)
print("Sample data")
head(bvit)[15:17]
print("Column names")
names(bvit)
```

## MetaboIndicators: metabolite sums and ratios
```{r, echo=FALSE}
ExpData(data = metabolInd, type = 1)
print("Sample data")
head(metabolInd)[15:17]
print("Sample column names")
names(metabolInd)[10:50]
```

## Sapient: untargeted metabolomics
```{r, echo=FALSE}
ExpData(data = sapient, type = 1)
print("Sample data")
head(sapient)[15:17]
print("Sample column names")
names(sapient)[50000:50100]
```

## Biocrates: info on each metabolite, provided by biocrates
```{r, echo=FALSE}
ExpData(data = biocrates, type = 1)
print("Sample data")
head(biocrates)[12:15]
print("Sample column names")
names(biocrates)[1:100]
```

## Biocrates normalized: normalized between plates
```{r, echo=FALSE}
ExpData(data = biocratesNorm, type = 1)
print("Sample data")
head(biocratesNorm)[12:15]
print("Column names are identical to the non-normalized biocrates dataset.")
```

# VITAL
```{r, echo=FALSE}
hmo <- read.csv("/data/KI/imic/data/raw_lab_data/vital/milk_analytes/HMO_VITAL.csv")
sapient <- read.csv("/data/KI/imic/data/raw_lab_data/vital/milk_analytes/Sapient_VITAL.csv")
bvit <- read.csv("/data/KI/imic/data/raw_lab_data/vital/milk_analytes/Allen_Bvit_VITAL.csv")
biocrates <- read.csv("/data/KI/imic/data/raw_lab_data/vital/milk_analytes/Biocrates_VITAL_NOT_Normalized.csv")
biocratesNorm <- read.csv("/data/KI/imic/data/raw_lab_data/vital/milk_analytes/Biocrates_VITAL_Normalized_toALL_NS.csv")
pbl <- read.csv("/data/KI/imic/data/raw_lab_data/vital/milk_analytes/PBL_VITAL_NOT_Normalized.csv")
pblNorm <- read.csv("/data/KI/imic/data/raw_lab_data/vital/milk_analytes/PBL_VITAL_Normalized.csv")
metabolInd <- read.csv("/data/KI/imic/data/raw_lab_data/vital/milk_analytes/MetaboIndicators/Biocrates_VITAL_Normalized_MetaboINDICATOR.csv")
```

## HMOs
Note: “Secretor”, “Diversity”, “Evenness”, “SUM_nmol/mL”, “SUM_ug/mL”, “Sia_nmol/mL”, and “Fuc_nmol/mL” are summary variables. The suffix “_nmol/mL”, “_ug/mL” and “_pct” indicate different units for the same HMOs.
```{r, echo=FALSE}
# hmo $ X = NULL
# hmoVital <- table1(~ ., data = hmo)
# saveRDS(hmoVital, file = paste0(BV_dir, "/results/hmoVital.RDS"))
# readRDS("/data/KI/imic/results/hmoVital.RDS")
ExpData(data = hmo, type = 1)
print("Sample data")
head(hmo)[15:17]
print("Sample column names")
names(hmo)[10:50]
```

## B-Vitamins
```{r, echo=FALSE}
# bvit $ X = NULL
# bvitVital <- table1(~ ., data = bvit)
# saveRDS(bvitVital, file = paste0(BV_dir, "/results/bvitVital.RDS"))
# readRDS("/data/KI/imic/results/bvitVital.RDS")
ExpData(data = bvit, type = 1)
print("Sample data")
head(bvit)[15:17]
print("Column names")
names(bvit)
```

## MetaboIndicators: metabolite sums and ratios
```{r, echo=FALSE}
ExpData(data = metabolInd, type = 1)
print("Sample data")
head(metabolInd)[15:17]
print("Sample column names")
names(metabolInd)[10:50]
```

## Sapient: untargeted metabolomics
```{r, echo=FALSE}
ExpData(data = sapient, type = 1)
print("Sample data")
head(sapient)[15:17]
print("Sample column names")
names(sapient)[50000:50100]
```

## Biocrates: info on each metabolite, provided by biocrates
```{r, echo=FALSE}
ExpData(data = biocrates, type = 1)
print("Sample data")
head(biocrates)[12:15]
print("Sample column names")
names(biocrates)[1:100]
```

## Biocrates normalized: normalized between plates
```{r, echo=FALSE}
ExpData(data = biocratesNorm, type = 1)
print("Sample data")
head(biocratesNorm)[12:15]
print("Column names are identical to the non-normalized biocrates dataset.")
```

## PBL: LC-MS proteomic data
```{r, echo=FALSE}
ExpData(data = pbl, type = 1)
print("Sample data")
head(pbl)[12:15]
print("Sample column names")
names(pbl)[1:100]
```

## PBL normalized
```{r, echo=FALSE}
ExpData(data = pblNorm, type = 1)
print("Sample data")
head(pblNorm)[12:15]
print("Column names are identical to the non-normalized PBL dataset.")
```











